Contrasting snow algae communities along an elevational gradient

### 2019-11-16 manuscript/manuscript.Rmd
minor edits

Snow algae blooms can color snow red, orange, or green, and can one or contain multiple species of algae. Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophyceae [@Remias2013;@Remias2019] have been reported. Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin responsible for the red color of some snow algae blooms. Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019c;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975]. However, some species are known from only one or two samples, and the recently described *Chlainomonas* and *Sanguina* are poorly explored, with only two species in each.

@@

Despite over a century of study, no one has successfully cultured the snow algae that produce red secondary pigments. In many green algae cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Shah2009;@Salome2019]; thus red snow algae could originate from green cells. Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow. Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashoma2016]; however it remains an open question whether green snow can develop into red snow.

@@

The obvious question is: what environmental factors driving those differences in community composition? Further research will be needed to determine the specific environmental factors driving community differences, but our field observations does provide some testable hypotheses. In forested areas, green cells would be shaded from damaging solar radiation, while above treeline pigmented cells of *Sanguina* and *Chloromonas* are likely protected by their astaxanthin pigment [@Gorton2003;Holzinger2006]. This fits with our observation that green cells of *Chloromonas krienitzii* were hidden below the snow surface, and pigmented cells were more prevalent on the surface (Fig. 4); possibly pigment production is triggered as snow level melts lower, and cells are exposed to higher light at the surface. However, this does not explain why *Chloromonas krienitzii* was not found at higher elevations: our *C. krienitzii* dominated samples were taken from sites that were ~100 m from the nearest tree, and were not notably shaded by topography in any way. Anecdotally, we observed *C. krienitzii* developed over running water, so their distribution could be linked to the presence of streams below the snowpack.

@@

We used light microscopy to characterize the cell morphologies in each sample. We prepared a slide of cells fixed in 2% gluteraldehyde, and at 400x magnification starting from the center of the slide counted up to 100 cells. We classified each cell as either *Sanguina nivaloides* [@Prochazkova2019a], *Chloromonas cf. nivalis* [@Matsuzaki2019;@Prochazkova2017], *Chloromonas cf. brevispina* [@Nedbalova2008, @Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "Other" based on similarity to published photographs.

>>>>

Snow algae blooms can color snow red, orange, or green, and can one or contain multiple species of algae. Most snow algae blooms are caused by green algae of class Chlorophyceae [@Leya2013], although blooms caused by Trebouxiophyceae [@Lutz2015] or Chrysophyceae [@Remias2013;@Remias2019] have been reported. Among the Chlorophyceae, both *Sanguina nivaloides* [@Prochazkova2019a] or *Chlainomonas* [@Prochazkova2018c] produce the red pigment astaxanthin responsible for the red color of some snow algae blooms. Orange snow is reported to be caused by *Sanguina aurentia* [@Prochazkova2019a] and some species of *Chloromonas* [@Prochazkova2019a;@Remias2013a], while algae lacking secondary pigments color the snow green, such as *Chloromonas pichinchae* [@Hoham1975b]. However, some species are known from only one or two samples, and the recently described *Chlainomonas* and *Sanguina* are poorly explored, with only two species in each.

@@

Despite over a century of study, no one has successfully cultured the snow algae that produce red secondary pigments. In model organism green algae *Chlamydomonas reinhardtii* cells produce secondary pigments in just one life stage, remaining green for much of their life cycle [@Salome2019], thus red snow algae could originate from green cells. Green snow is frequently found in waterlogged snow, sometimes alongside red snow, leading some to hypothesize that green snow could be an earlier life stage of red snow. Two studies comparing green and red snow using amplicon high-throughput sequencing (HTS) found they were caused by different species [@Lutz2015;@Terashima2017]; however it remains an open question whether green snow can develop into red snow.

@@

The obvious question is: what environmental factors driving those differences in community composition? Further research will be needed to determine the specific environmental factors driving community differences, but our field observations does provide some testable hypotheses. In forested areas, green cells would be shaded from damaging solar radiation, while above treeline pigmented cells of *Sanguina* and *Chloromonas* are likely protected by their astaxanthin pigment [@Gorton2003;@Holzinger2006]. This fits with our observation that green cells of *Chloromonas krienitzii* were hidden below the snow surface, and pigmented cells were more prevalent on the surface (Fig. 4); possibly pigment production is triggered as snow level melts lower, and cells are exposed to higher light at the surface. However, this does not explain why *Chloromonas krienitzii* was not found at higher elevations: our *C. krienitzii* dominated samples were taken from sites that were ~100 m from the nearest tree, and were not notably shaded by topography in any way. Anecdotally, we observed *C. krienitzii* developed over running water, so their distribution could be linked to the presence of streams below the snowpack.

@@

We used light microscopy to characterize the cell morphologies in each sample. We prepared a slide of cells fixed in 2% gluteraldehyde, and at 400x magnification starting from the center of the slide counted up to 100 cells. We classified each cell as either *Sanguina nivaloides* [@Prochazkova2019a], *Chloromonas cf. nivalis* [@Matsuzaki2019;@Prochazkova2017], *Chloromonas cf. brevispina* [@Nedbalova2008a;@Matsuzaki2015], *Chloromonas krienitzii* [@Matsuzaki2015] *Chlainomonas rubra* [@Novis2008], or "Other" based on similarity to published photographs.
